---
title: "Portland Public Schools Math Proficiency Report"
format: html
execute: 
  echo: false
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(here)
library(flextable)
library(gt)
library(scales)
```

```{r}
third_grade_math_proficiency_wide <-
  read_rds(here("data/third_grade_math_proficiency_dichotomous.rds")) |>
  filter(district == "Portland SD 1J") |>
  filter(
    school %in%
      c(
        "Abernethy Elementary School",
        "Ainsworth Elementary School",
        "Alameda Elementary School",
        "Arleta Elementary School",
        "Atkinson Elementary School"
      )
  ) |>
  select(year, school, percent_proficient) |>
  arrange(school) |>
  pivot_wider(
    id_cols = school,
    names_from = year,
    values_from = percent_proficient
  )
```

![Portland Public Schools](portland-public-schools-logo.svg){width=300px fig-align="center" fig-alt="Portland Public Schools logo"}

# Introduction

This is a report on math proficiency results in [Portland Public Schools (PPS)](https://www.pps.net/). The PPS mission statement is as follows:

> We provide rigorous, high-quality academic learning experiences that are inclusive and joyful. We disrupt racial inequities to create vibrant environments for every student to demonstrate excellence.^[https://www.pps.net/about/portland-public-schools-information/overview]

# Don’t Use the Default Output

```{r}
third_grade_math_proficiency_wide
```

# Flextable

```{r}
#| tbl-cap: Math proficiency among third graders in five Portland schools

third_grade_math_proficiency_wide |>
  mutate(`2018-2019` = percent(`2018-2019`)) |>
  mutate(`2021-2022` = percent(`2021-2022`)) |>
  flextable() |>
  set_header_labels(school = "School") |>
  # align(j = 2, align = "center") |>
  # width(j = 2, 4, unit = "cm") |>
  # width(j = 3, 2, unit = "cm")
  autofit()
```

# gt

```{r}
#| tbl-cap: Math proficiency among third graders in five Portland schools

third_grade_math_proficiency_wide |>
  gt() |>
  cols_label(school = "School") |>
  cols_width(
    school ~ px(100)
  ) |>
  # cols_align(
  #   columns = `2018-2019`,
  #   align = "center"
  # ) |>
  fmt_percent(
    columns = 2:3,
    decimals = 1
  )
```

# gt interactive

```{r}
third_grade_math_proficiency_wide_full <-
  read_rds(here("data/third_grade_math_proficiency_dichotomous.rds")) |>
  filter(district == "Portland SD 1J") |>
  select(year, school, percent_proficient) |>
  arrange(school) |>
  pivot_wider(
    id_cols = school,
    names_from = year,
    values_from = percent_proficient
  )
```

```{r}
third_grade_math_proficiency_wide_full |>
  gt() |>
  cols_label(school = "School") |>
  fmt_percent(
    columns = 2:3,
    decimals = 0
  ) |>
  opt_interactive(
    use_search = TRUE,
    use_highlight = TRUE
  )
```
